---
title: "Phosphorus Kinetics in DMSTAr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Phosphorus Kinetics in DMSTAr}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup,echo = FALSE}
library(DMSTAr)
```

## Summary

This vignette documents the DMSTA-style phosphorus kinetics available in DMSTAr: how kinetic parameters are constructed, what the coefficients mean, how tank-in-series settings interact with kinetics, and how to customize kinetics in a reproducible R workflow.

This vignette focuses on *within-case* phosphorus dynamics. Network routing and hydrology setup are covered in `vignette("Getting Started with DMSTAr")`.

## Introduction

DMSTAr implements a modular version of DMSTA-style phosphorus dynamics. The goal is to keep the scientific intent clear (what the parameters mean) while supporting extensible implementation (how users provide parameters).

## Background

### Conceptual Structure

DMSTA represents phosphorus cycling with two storages:

-   water-column concentration/state (C)

-   labile biomass/sediment storage (S)

Three internal fluxes are represented with rate coefficients:

-   removal (K1), recycle (K2), burial (K3)

Core internal flux forms (conceptually) are:

-   $Removal \propto K1 \cdot F_{c}(C) \cdot F_{z}(z) \cdot C \cdot S$

-   $Recycle \propto K2 \cdot S^{2}$

-   $Burial \propto K3 \cdot S$

Where

-   $K1$, $K2$, $K3$ = removal, recycling and burial rate coefficients

-   C = water-column P concentration

-   S = liable P Storage

-   $F_{c}$ and $F_{z}$ = efficiency modifiers

#### Concentration-dependent scale factor ($F_{c}$)

$F_{c}$ limits removal efficiency at low water column P concentration. DMSTA, at its core, assumes that biological and physicochemical uptake processes become less efficient as C approaches background levels. Therefore, $F_{c}$ represents declining phosphorus uptake efficiency as water‑column concentrations approach low‑P conditions. Ultimately, this factor prevents the model from unrealistically removing phosphorus at full strength when concentrations are already very low.

In practice, $F_{c}$ is governed by the parameter "C2" also called Chalf. It is defined as the concentration at which removal efficiency is \~50% of the maximum. Conceptually, at high concentrations, $F_{c} \approx 1$, whereas at low concentrations, $F_{c}$ is effectively 0.

<br>

```{r Fc calc, echo = FALSE,fig.align = "center",fig.cap = "Graphical representation of $F_c$.",fig.width = 6.5,fig.height = 4}
## DMSTA uses Michaelis- Menten / half saturation


Fc <- function(C, Chalf) {
  C / (C + Chalf)
}


C  <- seq(0, 200, by = 10)  # TP concentration (µg/L)
Chalf <- 50                 # typical DMSTA value
Fc_dat <- data.frame(C = C,
                     Fc = Fc(C, Chalf))
par(mar=c(2,3,1,1.5),oma = c(2,2,0.5,0.5))
ylim.val <- c(0,1);by.y <- 0.2;ymaj <- seq(ylim.val[1],ylim.val[2],by.y)
xlim.val <- c(0,200);by.x <- 50;xmaj <- seq(xlim.val[1],xlim.val[2],by.x)

plot(Fc~C, Fc_dat,type = "n",ylim=ylim.val,xlim=xlim.val,las=1,ann=F)
abline(h=ymaj,v=xmaj,lty=3,col="grey",lwd=0.5)
lines(Fc~C, Fc_dat,lwd=2)
abline(h = c(0, 1), lty = 2,lwd=2,col="red")
mtext(side=3,adj=0.01,expression("Concentration-dependent scale factor " * F[c]))
mtext(side=1,line=2.5,expression("Water-column TP conc. ("*mu*"g"~"L"^"-1"*")"))
mtext(side=2,line=2.75,expression(F[c]),cex=1.25)
```


<br>

#### Depth Dependent scale factor ($F_{z}$)

$F_{z}$ adjusts removal efficiency based on water depth. Phosphorus removal processes in STAs are most effective over a preferred depth range. Very shallow or very deep water reduces contact with vegetation, periphyton, and sediments. $F_z$ encodes that assumption without adding new state variables accounting for the reducted P removal efficiency below optimal water depths.

In DMSTA (and `DMSTAr`), $F_z$ is controlled by three depth parameters, depth of maximum or saturated uptake ($Z1$), lower penalty depth ($Z2$) and upper penalty depth ($Z3$). Conceptually, depth near $Z1$, $F_z \approx 1$. However, when depth is $<Z2$ or $>Z3$ $F_z <1$.

<br>

```{r Fz calc, echo = FALSE,fig.align = "center",fig.cap = "Graphical representation of $F_z$.",fig.width = 6.5,fig.height = 4}
Fz <- function(z, Z1, Z2, Z3) {
  # Based on DMSTA2 calibration
  ifelse(
    z < Z1, z / Z1,
    ifelse(
      z <= Z2, 1,
      ifelse(
        z < Z3, (Z3 - z) / (Z3 - Z2),
        0
      )
    )
  )
}

z  <- seq(0, 250, by = 1)  # water depth (cm)

Z1 <- 40    # saturated uptake depth
Z2 <- 100   # lower penalty depth
Z3 <- 200   # upper penalty depth

Fz_dat <- data.frame(Z = z,
                     Fz = Fz(z, Z1, Z2, Z3))

par(mar=c(2,3,1,1.5),oma = c(2,2,0.5,0.5))
ylim.val <- c(0,1);by.y <- 0.2;ymaj <- seq(ylim.val[1],ylim.val[2],by.y)
xlim.val <- c(0,200);by.x <- 50;xmaj <- seq(xlim.val[1],xlim.val[2],by.x)

plot(Fz~Z, Fz_dat,type = "n",ylim=ylim.val,xlim=xlim.val,las=1,ann=F)
abline(h=ymaj,v=xmaj,lty=3,col="grey",lwd=0.5)
lines(Fz~Z, Fz_dat, lwd=2)
abline(h = c(0, 1), lty = 2,lwd=2,col="red")
mtext(side=3,adj=0.01,expression("Depth-dependent scale factor " * F[z]))
mtext(side=1,line=2.5,"Water Depth (cm)")
mtext(side=2,line=2.75,expression(F[z]),cex=1.25)

# Fz_dat$Fz_EAV <- with(Fz_dat,pmin(1,Z/40)); #DMSTA1 
# lines(Fz_EAV~Z, Fz_dat,lty=2 ,lwd=2,col="blue")
```

<br>

<!-- ### Tanks-in-series -->

<!-- Tanks-in-series (TIS) also referred to as completely stirred tank reactor (CSTR) is a series of complete-mix reactors in series. These models can be used to capture important features of wetland detention time distribution. -->

<!-- Within a cell, the model is solved across a tanks-in-series representation. Outflow from tank `i` becomes inflow to tank `i+1`; the final tank represents the cell discharge concentration. -->

## Parameterization

In `DMSTAr`, phosphorus kinetics are created using a model-type builder (e.g. "STA", "PSTA", "RES"). Builders return a standardized set of raw parameters (including at least `C1000`, `Cstar`, `Ks`), and DMSTAr derives $K1$, $K2$ and $K3$ plus a model indicator PModel.

`DMSTAr`, consistent with DMSTA, supports two internal formulations

-   PModel 1: standard model

-   PModel 2: special case, triggered when `Cstar` \< 0

Phosphorus kinetics are calculated from a series of functions, some internal to the package but can be called using `DMSTAr:::`. Here is an example of a function to calculate the various K values and identifies which PModel is used in downstream analytical processes

```{r Pmodel1}
DMSTAr:::compute_DMSTA_kvals(C1000 = 22, Cstar = 3, Ks = 16)

```

When `Cstar` \< 0, PModel 2 methodology is used to estimate K value

```{r PModel2}
DMSTAr:::compute_DMSTA_kvals(C1000 = 22, Cstar = -3, Ks = 16)

```

Similar to DMSTA, `DMSTAr` calculates kinetic variables into three slots to perform P simulations depending on what models are applied. To build the necessary kinetics, `build_P_kin_slots` is used where parameters and model types are supplied. Currently the default models are `STA`, `PSTA` and `RES` but there is functionality to change the model types in the three slot configuration. Currently the function includes a `Dpy` argument which is a factor to convert per day to pre year. It is possible in the this argument will be moved into the function as a default.

```{r kinetics}
pparams <- list(
  # shared / STA
  C1000 = 22, Cstar = 3, Ks_per_yr = 16,
  Z1 = 40, Z2 = 100, Z3 = 200,
  K2Coef1 = 0.1, Chalf = 50, SeasonalFactor = 1,
  # PSTA
  Ytrans = 1, Ysigma = 1, C1000_2 = 50, ks_2 = 20, zh_2 = 10,
  # RES
  k_depth_penalty = 0.5,
  DutyCycle = 0.95
)

kin_out <- build_P_kin_slots(
  mods = c("STA", "PSTA", "RES"),
  pparams = pparams,
  Dpy = 365.25
)
kin_out
```
